---
title: 'Roots and Resilience: How Family Adversity Shapes Child and Adult Outcomes'
author: "Maliha"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

Introduction

This project explores how adverse childhood experiences (ACEs) — 
particularly parental depression and household dysfunction — shape 
child and adult outcomes, and what factors predict resilience.

Setup

```{r setup, message=FALSE, warning=FALSE}
install.packages(c("tidyverse", "haven", "skimr", "ggplot2", "shiny"))
library(tidyverse)
library(skimr)
library(ggplot2)
```


 Load Data

```{r load-data}
install.packages("medicaldata")
library(medicaldata)

data(package = "medicaldata")
```

```{r explore-datasets}
data(package = "medicaldata")$results[, c("Item", "Title")]
```

```{r load-aces}
remove.packages("medicaldata")

install.packages("NHANES")
library(NHANES)

data(NHANES)
glimpse(NHANES)
```


What socioeconomic and behavioral factors predict depression severity in adults — and what looks protective?

First Look at Our Key Variables

```{r explore-key-vars}
NHANES %>%
  select(Age, Gender, Poverty, Depressed, DaysMentHlthBad, 
         LittleInterest, SleepTrouble, PhysActive, HHIncomeMid) %>%
  skim()
```

Depression Distribution

```{r depression-viz}
NHANES %>%
  filter(!is.na(Depressed)) %>%
  ggplot(aes(x = Depressed, fill = Depressed)) +
  geom_bar() +
  labs(
    title = "Depression Frequency in NHANES Sample",
    x = "Depressed (Days in past 2 weeks)",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Majority report no depression — expected
But ~1000 report "Several" days and ~450 report "Most" days
That's roughly 1 in 7 people in this sample experiencing significant depressive symptoms

That's not a small signal. That's a public health pattern.
Adding poverty:
Depression by Poverty Level

```{r depression-poverty}
NHANES %>%
  filter(!is.na(Depressed), !is.na(Poverty)) %>%
  ggplot(aes(x = Depressed, y = Poverty, fill = Depressed)) +
  geom_boxplot() +
  labs(
    title = "Poverty Level by Depression Severity",
    x = "Depression Frequency",
    y = "Poverty Index (lower = more poverty)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```


What this shows:

"None" group — median poverty index ~3.3 (more resources)
"Several" group — median drops to ~2.3
"Most" group — median drops further to ~1.4 (closest to poverty line)

The pattern is clear and consistent — as depression severity increases, poverty index decreases. More depression, fewer resources. Every step of the gradient moves in the same direction.
That's not noise. That's a signal worth writing about.

## Finding 1: Poverty and Depression Move Together

Analysis of the NHANES sample reveals a consistent gradient between 
economic hardship and depression severity. Individuals reporting 
depressive symptoms "most" days show markedly lower poverty index 
scores compared to those reporting none, suggesting that financial 
stress and mental health burden are deeply intertwined — not independent 
risk factors.

```{r depression-sleep}
NHANES %>%
  filter(!is.na(Depressed), !is.na(SleepTrouble)) %>%
  ggplot(aes(x = Depressed, fill = SleepTrouble)) +
  geom_bar(position = "fill") +
  labs(
    title = "Sleep Trouble by Depression Severity",
    x = "Depression Frequency",
    y = "Proportion",
    fill = "Sleep Trouble"
  ) +
  theme_minimal()
```

What this shows:

"None" group — ~20% report sleep trouble
"Several" group — jumps to ~38%
"Most" group — crosses 50%. More than half report sleep trouble

The gradient holds again. Every step up in depression severity brings more sleep disruption. And that 50% threshold in the "Most" group is striking — sleep trouble stops being a symptom and starts looking like a co-condition.

## Finding 2: Sleep Disruption Scales with Depression Severity

Among those reporting depressive symptoms most days, over half 
also report trouble sleeping — more than double the rate seen in 
those with no depression. This pattern suggests sleep disruption 
is not merely a side effect but a compounding factor in severe 
depression, potentially creating a reinforcing cycle of poor 
mental health and physical dysregulation.

```{r depression-physactive}
NHANES %>%
  filter(!is.na(Depressed), !is.na(PhysActive)) %>%
  ggplot(aes(x = Depressed, fill = PhysActive)) +
  geom_bar(position = "fill") +
  labs(
    title = "Physical Activity by Depression Severity",
    x = "Depression Frequency",
    y = "Proportion",
    fill = "Physically Active"
  ) +
  theme_minimal()
```

first resilience finding.
What this shows:

"None" group — ~58% are physically active
"Several" group — drops to ~47%
"Most" group — drops to ~34%

The gradient flips. As depression gets worse, physical activity decreases. Or framed the other way — physical activity is consistently more present in the less depressed group.
This is your first answer to what gets people out — or at least what looks different in people who aren't stuck.

## Finding 3: Physical Activity as a Protective Factor

Physical activity follows a clear inverse relationship with depression 
severity. Nearly 6 in 10 people reporting no depression are physically 
active, compared to only 1 in 3 among those experiencing depressive 
symptoms most days. While causality cannot be established from this 
data alone, the pattern is consistent with research suggesting movement 
as a meaningful protective factor against depression.


Finding 1: Poverty and depression move together
Finding 2: Sleep disruption scales with severity
Finding 3: Physical activity is protective

## Gender and Depression

```{r depression-gender}
NHANES %>%
  filter(!is.na(Depressed), !is.na(Gender)) %>%
  ggplot(aes(x = Depressed, fill = Gender)) +
  geom_bar(position = "fill") +
  labs(
    title = "Depression Severity by Gender",
    x = "Depression Frequency",
    y = "Proportion",
    fill = "Gender"
  ) +
  theme_minimal()
```

What this shows:

"None" group — roughly 50/50 male/female, slightly more male
"Several" group — shifts to ~55% female
"Most" group — shifts further to ~60% female

As depression severity increases, women are increasingly overrepresented. Men don't disappear from the severe group — but the gender gap widens at every step.

## Finding 4: Gender and Depression Severity

Women are disproportionately represented among those experiencing 
the most severe depressive symptoms. While the sample is roughly 
equal by gender overall, women account for nearly 60% of those 
reporting depressive feelings most days. This pattern is consistent 
with well-documented gender disparities in depression prevalence 
and raises important questions about the structural, biological, 
and social factors that place women at higher risk — including 
caregiving burden, economic inequality, and hormonal vulnerability.


the seed for Project 2:
The biological angle is genuinely rich territory. Possible question:
"How do biological markers — cortisol, inflammation, hormonal patterns — differ across depression severity, and do they interact with gender?"
NHANES actually has some biological variables already — testosterone, cholesterol. 
for deeper biology:

NHANES lab files — includes CRP (inflammation marker), cortisol proxies
UK Biobank — massive, includes genetics, biological assays, mental health
ABCD Study — brain imaging + family environment + child outcomes

4 findings:

Poverty and depression move together
Sleep disruption scales with severity
Physical activity is protective
Women are overrepresented in severe depression

# Conclusion

```{r conclusion}
# Summary of key findings
```

## What This Analysis Shows

This exploratory analysis of NHANES data reveals a consistent, 
multi-layered portrait of depression in the US adult population. 
Economic hardship, sleep disruption, physical inactivity, and 
gender each independently track with depression severity — and 
likely interact with one another in ways this cross-sectional 
data cannot fully capture.

## What This Analysis Cannot Tell Us

Causality. This data is a snapshot, not a timeline. We cannot 
say poverty *causes* depression, or that inactivity *leads to* 
worse outcomes — only that they co-occur in a meaningful pattern. 
Longitudinal data would be required to establish direction.

## What Comes Next

Project 2 will examine biological markers — inflammation, 
hormonal patterns, and physiological stress responses — and 
their relationship to depression severity and gender. Using 
NHANES lab files and potentially UK Biobank data, the goal 
is to move beyond behavioral correlates into the biological 
substrates of what keeps people stuck.


## Summary: How All Factors Relate

```{r correlation-summary}
library(reshape2)

# Create numeric version of key variables
nhanes_numeric <- NHANES %>%
  filter(!is.na(Depressed), !is.na(Poverty), 
         !is.na(SleepTrouble), !is.na(PhysActive),
         !is.na(Gender)) %>%
  mutate(
    Depressed_n = as.numeric(Depressed),
    SleepTrouble_n = ifelse(SleepTrouble == "Yes", 1, 0),
    PhysActive_n = ifelse(PhysActive == "Yes", 1, 0),
    Gender_n = ifelse(Gender == "female", 1, 0)
  ) %>%
  select(
    Depression = Depressed_n,
    Poverty = Poverty,
    Sleep_Trouble = SleepTrouble_n,
    Physical_Activity = PhysActive_n,
    Gender_Female = Gender_n,
    Age = Age,
    HH_Income = HHIncomeMid
  )

# Correlation matrix
cor_matrix <- cor(nhanes_numeric, use = "complete.obs")

# Melt and plot
melted_cor <- melt(cor_matrix)

ggplot(melted_cor, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "steelblue", mid = "white", high = "tomato",
    midpoint = 0, limit = c(-1, 1),
    name = "Correlation"
  ) +
  geom_text(aes(label = round(value, 2)), size = 3) +
  labs(
    title = "Correlation Matrix: Depression and Associated Factors",
    x = "", y = ""
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
What to read here:
Depression row (bottom):

Sleep_Trouble: +0.21 — strongest positive relationship. More sleep trouble, more depression
Poverty: -0.19 — lower poverty index (more poverty), more depression
HH_Income: -0.18 — lower income, more depression
Physical_Activity: -0.11 — less activity, more depression
Gender_Female: +0.07 — small but present female skew
Age: -0.02 — essentially no relationship

The bonus finding — Poverty and HH_Income: +0.88
That's expected — they measure similar things. Actually validates your data is clean.

## Finding 5: Summary — What Matters Most

The correlation matrix confirms and quantifies the patterns observed 
throughout this analysis. Sleep trouble shows the strongest individual 
relationship with depression (r = 0.21), followed closely by poverty 
(r = -0.19) and household income (r = -0.18). Physical activity shows 
a modest protective relationship (r = -0.11). Gender shows a small 
but consistent female skew (r = 0.07).

Notably, no single factor dominates. Depression appears to be 
distributed across multiple compounding burdens — economic, 
physiological, and behavioral — rather than driven by any one cause. 
This has direct implications for intervention: addressing only one 
factor may be insufficient.


